Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex
Add samples per day supplementary figure

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model \cite{Li2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

Sc2ts uses collection dates to build a time-resolved ARG.
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions),
likely due to errors in their collection dates
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1,
which is estimated to have emerged in November, 2021; REF).
To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”.
We define the “HMM cost” of a sample to be
the total penalty cost for switches and mismatches (mutations) in its copying path.
We parameterize the LS model such that each mismatch has a penalty cost of one
and a switch has a penalty cost of $k$ (here, set to 4),
which determines how much mutation is favored over recombination when accounting for a sample (\ref{fig:methodology}).
If the HMM cost of a sample exceeds a threshold (here, set to 7),
sc2ts does not immediately add the sample to the ARG.

This approach resulted in preliminary ARGs that had noticeably fewer time travelers,
but we found that it also excluded the major “saltation” events that characterize many VOCs.
These saltation lineages are characterized by a high number of accumulated mutations,
and are hypothesized to emerge from chronic infections
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}.
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron).
To distinguish time travelers with erroneous dates from true saltation lineages,
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
If on subsequent days these lineages form a sufficiently large group of samples,
sc2ts adds the group back to the ARG.
This “retrospective sample grouping” successfully captured the emergence of most saltation lineages (Figure SX).
The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
in which specific "seed" samples were identified for unconditional inclusion
(for details of how seed samples were chosen, see STAR Methods).
First arising in geographic regions with low sampling density,
the initial origins of these VOC are challenging to infer (REF).

@@

\section{Major deletions in the ARG}
% Referenced by the main Results.
Basically, we want to say we did a lot of work to integrate signals of major
indels into the ARG, making it an even more powerful resource overall. Also,
indels are encoded by the Mutation-Annotated Tree format used by UShER
phylogenies (though PanMAT does this).

@@

\section{Identification of robust recombination events}
Before characterizing recombination events in detail, we sought to remove
probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
sequencing protocols have been documented (Turakhia et al., 2022), which can
confound recombination detection (De Maio et al., 2024). By inspecting copying
patterns at sites that differ between recombinant parents, we find a convincing
signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized as having fewer than four
such loci supporting the left or right parent on either side of a suggested
breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
measure identifies 543 as artifactual, leaving 386 “robust” events (Figure
S1A). The copying patterns depicted in Figure S1B illustrate the genomic
evidence for three recombination events identified as artifactual, compared to
a marginally robust one.

@@

\section*{Supplementary Figures}

>>>>

Sc2ts reconstructs an ARG by incrementally adding batches of sequences as they become available
(Figure~\ref{fig:methodology}; for details, see STAR Methods).
First, sc2ts infers likely ``copying paths'' connecting each sample in the batch to node(s) in the current ARG
(allowing recombination) using the Li and Stephens (LS) model\cite{Li2003},
which is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic inheritance.
As recombination is relatively rare in viruses,
we would expect most samples to copy from a single parent in the ARG.
Recombinant samples, however, copy from two or more parents, with switching (recombination) in the copying path
between the parents resulting in different genomic segments inherited from the different parents.
Next, for clusters within the batch that are inferred to copy from the same ARG node,
sc2ts uses standard phylogenetic techniques to resolve relationships within the cluster.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced by the inherent greediness of this strategy.

Sc2ts uses collection dates to build a time-resolved ARG.
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions),
likely due to errors in their collection dates
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1,
which is estimated to have emerged in November, 2021; REF).
To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”.
We define the “HMM cost” of a sample to be
the total penalty cost for switches and mismatches (mutations) in its copying path.
We parameterize the LS model such that each mismatch has a penalty cost of one
and a switch has a penalty cost of $k$ (here, set to 4),
which determines how much mutation is favored over recombination when accounting for a 
sample (Figure~\ref{fig:methodology}).
If the HMM cost of a sample exceeds a threshold (here, set to 7),
sc2ts does not immediately add the sample to the ARG.

This approach resulted in preliminary ARGs that had noticeably fewer time travelers,
but we found that it also excluded the major ``saltation'' events that characterize many VOCs.
These saltation lineages are characterized by a high number of accumulated mutations,
and are hypothesized to emerge from chronic infections
during which the virus is exposed to individual-specific immune selective pressures \cite{Markov2023}.
Though uncommon, such lineages drove large epidemic waves during the pandemic (notably Alpha, Delta, and Omicron).
To distinguish time travelers with erroneous dates from true saltation lineages,
we reevaluate the evidence for each lineage with a high HMM cost as additional samples are added.
If on subsequent days these lineages form a sufficiently large group of samples,
sc2ts adds the group back to the ARG.
This ``retrospective sample grouping'' successfully captured the emergence of
most saltation lineages (Figure~\ref{fig:samples_per_day}).
The emergence of the Delta, BA.1, and BA.2 lineages required some manual intervention,
in which specific "seed" samples were identified for unconditional inclusion
(for details of how seed samples were chosen, see STAR Methods).
First arising in geographic regions with low sampling density,
the initial origins of these VOC are challenging to infer (REF).

@@

\section{Major deletions in the ARG}\label{sec_supplement_dels}
% Referenced by the main Results.
Basically, we want to say we did a lot of work to integrate signals of major
indels into the ARG, making it an even more powerful resource overall. Also,
indels are encoded by the Mutation-Annotated Tree format used by UShER
phylogenies (though PanMAT does this).

@@

\section{Identification of robust recombination
events}\label{sec_supplement_robust_recombs}
Before characterizing recombination events in detail, we sought to remove
probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
sequencing protocols have been documented (Turakhia et al., 2022), which can
confound recombination detection (De Maio et al., 2024). By inspecting copying
patterns at sites that differ between recombinant parents, we find a convincing
signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized as having fewer than four
such loci supporting the left or right parent on either side of a suggested
breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
measure identifies 543 as artifactual, leaving 386 “robust” events (Figure
S1A). The copying patterns depicted in Figure S1B illustrate the genomic
evidence for three recombination events identified as artifactual, compared to
a marginally robust one.

@@

\section*{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{samples-per-day}
    \caption{TODO caption}\label{fig:samples_per_day}
\end{figure}
